PanGraph: scalable bacterial pan-genome graph construction \\ Supplementary Information

### 2022-09-06 supplementary.tex
feat: continued supplementary

\section{Benchmark of \textit{PanGraph} on synthetic data}

@@

\begin{figure}[h]
    \includegraphics[width=.9\textwidth]{figs_suppl/benchmark_suppl.pdf}
    \caption{{\bf Computational performance of PanGraph algorightm on synthetic data, as a function of dataset size}. We run PanGraph on artificially generated datasets consisting of a variable number of isolates with genomes of variable length (colorscale). For each condition we display the average and standard deviation over 50 different runs of: (A) algorithm wall-time, (B) maximum memory requirement and (C) average percentage of cpu used on a total of 8 cores.}
    \label{fig:benchmark-perf-suppl}
\end{figure}

@@

\begin{figure}[h]
    \includegraphics[width=0.5\textwidth]{figs_suppl/snps_rate_vs_divergence.pdf}
    \caption{{\bf mutation rate vs divergence on synthetic data}. We compare the mutation rate $\mu$ of our simulations with the average pairwise divergence of sequences in pancontigs of the resulting pangenome graphs. This is done for different values of the rate of HGT $h$ and for three different alignment kernels. As expected, the divergence is only marginally influenced by the HGT rate, but the saturating value of the divergence depends strongly on the choice of alignment kernel, with \textit{mmseqs2} being able to merge sequences with average divergence higher than 10\%. To find the relationship between the mutation rate $\mu$ and the average pairwise divergence of the sequences we peform a linear fit on data points for $\mu \leq 0.002$. This provides the conversion factor $\avg{d} \sim 20.8 \, \mu$.}
    \label{fig:snps-suppl}
\end{figure}

@@

\begin{figure}[h]
    \includegraphics[width=\textwidth]{figs_suppl/accuracy_comparison.pdf}
    \caption{{\bf Accuracy of PanGraph algorithm on synthetic data}.}
    \label{fig:benchmark-accuracy-suppl}
\end{figure}

\section{Benchmark of \textit{PanGraph} on real data}

\begin{figure}[h]
    \includegraphics[width=\textwidth]{figs_suppl/panx_benchmark_suppl.pdf}
    \caption{{\bf Benchmark on real data}.}
    \label{fig:panx-benchmark-suppl}
\end{figure}

@@

Given a graph composed of many different genomes, PanGraph is capable of marginalizing this graph on a subset of strains with the \verb|marginalize| command. To test this feature we built pangenome graphs on genomes from 5 different species: \textit{Klebsiella Pneumoniae} (KP), \textit{Helicobacter Pylori} (HP), \textit{Prochlorococcus Marinus} (PM), \textit{Mycobacterium Tuberculosis} (MT) and \textit{Escherichia Coli} (EC), extracted from RefSeq.\footnote{The list of accession numbers for each species and pairs is available on the repository: \url{https://github.com/neherlab/pangraph/blob/feat/finalize-paper/script/config/projection_strains.json}} For each strain we took 50 different strains (10 for PM) and built an unique pangenome graph using the \verb|build| command with standard values of the parameters. For each species we then randomly picked 50 different pairs of strains (45 for PM) and extracted the corresponding sub-graphs using the \verb|marginalize| command. We compared each of these graphs to their counterpart generated simply by using the \verb|build| command on the corresponding pair of strains.\\
The comparison between the marginalized (MG) and pairwise (PW) graph consists in splitting each genome using the overlap of the breakpoints given by the two different graphs. Then for each resulting segment we check whether it is shared with the other strain of the pair on either of the graphs. This gives raise to four different possibilities, corresponding to the segment being shared (1) or not (0) on either of the two graphs (MG/PW). We consider four different classes of segments: \textit{agree on sharing} (1-1 or 0-0), \textit{shared on both} (1-1), \textit{private on both} (0-0) and \textit{disagree on sharing} (1-0 or 0-1). For each class we measure the total and average length of the segments belonging to that class, and display the results in Fig.~\ref{fig:marginalize-suppl}.\\
For all species except for PM we find that MG graphs are compatible with their PW counterpart: segments on which the graphs \textit{disagree on sharing} represent only a minor fraction of the genome (much smaller than the fraction of shared or private segments) and their size is compatible with the default block length threshold of PanGraph (100 bp). PM represents and exception, with the great majority of segments being private on each graph and having large size. This expected since for PM the strains considered are highly diverged, above the sensitivity of minimap2 ($\sim 27\%$ core divergence). This causes the majority of blocks not to be merged, and generates a strong order dependence. The size of segments with \textit{disagreement on sharing} is also the largest, some segments having size of 10s of kbp.

\begin{figure}[h]
    \includegraphics[width=.8\textwidth]{figs_suppl/proj_fig.pdf}
    \caption{{\bf Marginalized graphs are compatible with pairwise graphs}. Comparison of marginalized (MG) and pairwise (PW) built graphs for 5 different species. For each pair of strains we measure the length of segments in which MG and PW graphs agree on whether the segments are shared on both strains (\textit{agree on sharing}). These can either be both shared (\textit{shared on both}) or both private (\textit{private on both}). We also measure the length of segments on which the MG and PW graphs disagree (\textit{disagree on sharing}). For each of these for cases we report the total length as fraction of the average genome (left column) and the average size of each segment (right column). For each category we display the distribution of average results for 50 different pairs (45 for PM).
    }
    \label{fig:marginalize-suppl}
\end{figure}

>>>>

\section{Benchmark of PanGraph on synthetic data}

@@

\begin{figure}[htb]
    \includegraphics[width=.9\textwidth]{figs_suppl/benchmark_suppl.pdf}
    \caption{{\bf Computational performance of PanGraph algorightm on synthetic data, as a function of dataset size}. We run PanGraph on artificially generated datasets consisting of a variable number of isolates with genomes of variable length (colorscale). For each condition we display the average and standard deviation over 50 different runs of: (A) algorithm wall-time, (B) maximum memory requirement and (C) average percentage of cpu used on a total of 8 cores.}
    \label{fig:benchmark-perf-suppl}
\end{figure}

@@

\begin{figure}[htb]
    \includegraphics[width=0.5\textwidth]{figs_suppl/snps_rate_vs_divergence.pdf}
    \caption{{\bf mutation rate vs divergence on synthetic data}. We compare the mutation rate $\mu$ of our simulations with the average pairwise divergence of sequences in pancontigs of the resulting pangenome graphs. This is done for different values of the rate of HGT $h$ and for three different alignment kernels. As expected, the divergence is only marginally influenced by the HGT rate, but the saturating value of the divergence depends strongly on the choice of alignment kernel, with \textit{mmseqs2} being able to merge sequences with average divergence higher than 10\%. To find the relationship between the mutation rate $\mu$ and the average pairwise divergence of the sequences we peform a linear fit on data points for $\mu \leq 0.002$. This provides the conversion factor $\avg{d} \sim 20.8 \, \mu$.}
    \label{fig:snps-suppl}
\end{figure}

@@

\begin{figure}[htb]
    \includegraphics[width=\textwidth]{figs_suppl/accuracy_comparison.pdf}
    \caption{{\bf Accuracy of PanGraph algorithm on synthetic data}. For each of our generated datasets we evaluate the average displacement of inferred pancontig breakpoints. The figure represents the cumulative distribution of average misplacements, stratified by the average pairwise divergence of sequences in pancontigs (see colorbar on the right). Each column corresponds to a different alignment kernel used to infer the pangenome graph. As divergence increases we observe a transition in the precision of breakpoint position estimation. At low divergence the position of most breakpoints is correctly estimated withing a few bps, while at high divergence most breakpoints can be misplaced of multiple hundreds of bps. The transition value for the divergence depends on the alignemnt kernel used, with \textit{mmseqs2} having the highest accuracy at the expense of higher computational time.}
    \label{fig:benchmark-accuracy-suppl}
\end{figure}

\section{Benchmark of PanGraph on real data}

We benchmarked \textit{PanGraph} on real data, focusing on the computational performances and the properties of the resulting pangenome graphs.

\subsection{Dataset selection}

We downloaded from RefSeq \cite{o2016reference} complete chromosomes from 5 different bacterial species, listed in Table~\ref{table:panx-dataset}.\footnote{A complete list of accession numbers can be found in...}. These genomes had previously been analyzed with PanX \cite{ding2018panx}. For each species we also report the average genome lenght and average pairwise divergence on core genome.

% TODO: also report average core genome size?
% TODO: avoid talking about PanX?
% TODO: add link with accesion numbers and download?

\begin{table}[hb]
    \begin{tabular}{l l l l}
        \hline\hline
        Species         & N. isolates & Avg. chromosome length [Mbp] & Avg. pairwise divergence (core genes) \\
        \hline
        E.~Coli         & 307         & 4.98                         &                                       \\
        K.~Pneumoniae   & 109         & 5.35                         &                                       \\
        H.~Pylori       & 85          & 1.62                         &                                       \\
        M.~Tuberculosis & 51          & 4.40                         &                                       \\
        P.~Marinus      & 10          & 1.80                         &                                       \\
        \hline
    \end{tabular}
    \caption{{\bf Dataset overview}. We downloaded complete chromosome sequences from RefSeq for 5 different bacterial species. For each species we report the number of isolates downloaded, along with the average chromosome length and average pairwise divergence on core genes.}
    \label{table:panx-dataset}
\end{table}

\subsection{Pangenome graph construction}

We built pangenome graphs using five different options for the alignment kernel:
\begin{itemize}
    \item kernel \textit{minimap2} with option \textit{asm10} and default values for the energy parameters $\alpha=100$, $\beta=10$.
    \item kernel \textit{minimap2} with option \textit{asm20} and default values for the energy parameters $\alpha=100$, $\beta=10$.
    \item kernel \textit{mmseqs2} and default values for the energy parameters $\alpha=100$, $\beta=10$.
    \item kernel \textit{minimap2} with option \textit{asm10} and null energy parameters $\alpha=0$, $\beta=0$.
    \item kernel \textit{mmseqs2} and null energy parameters $\alpha=0$, $\beta=0$.
\end{itemize}
% TODO: describe expectations and reasons for the choice

\subsection{Benchmark results}

\begin{figure}[htb]
    \includegraphics[width=\textwidth]{figs_suppl/panx_benchmark_suppl.pdf}
    \caption{{\bf Benchmark on real data}.}
    \label{fig:panx-benchmark-suppl}
\end{figure}

@@

Given a graph composed of many different genomes, PanGraph is capable of marginalizing this graph on a subset of strains with the \verb|marginalize| command. To test this feature we built pangenome graphs on genomes from 5 different species: \textit{Klebsiella Pneumoniae} (KP), \textit{Helicobacter Pylori} (HP), \textit{Prochlorococcus Marinus} (PM), \textit{Mycobacterium Tuberculosis} (MT) and \textit{Escherichia Coli} (EC), extracted from RefSeq.\footnote{The list of selected accession numbers for each species and pair is available on the repository: \url{https://github.com/neherlab/pangraph/blob/feat/finalize-paper/script/config/projection_strains.json}} For each strain we took 50 different strains (10 for PM) and built an unique pangenome graph using the \verb|build| command with standard values of the parameters. For each species we then randomly picked 50 different pairs of strains (45 for PM) and extracted the corresponding sub-graphs using the \verb|marginalize| command. We compared each of these graphs to their counterpart generated simply by using the \verb|build| command on the corresponding pair of strains.\\
The comparison between the marginalized (MG) and pairwise (PW) graph consists in splitting each genome using the overlap of the breakpoints given by the two different graphs. Then for each resulting segment we check whether it is shared with the other strain of the pair on either of the graphs. This gives raise to four different possibilities, corresponding to the segment being shared (1) or not (0) on either of the two graphs (MG/PW). We consider four different classes of segments: \textit{agree on sharing} (1-1 or 0-0), \textit{shared on both} (1-1), \textit{private on both} (0-0) and \textit{disagree on sharing} (1-0 or 0-1). For each class we measure the total and average length of the segments belonging to that class, and display the results in Fig.~\ref{fig:marginalize-suppl}.\\
For all species except for PM we find that MG graphs are compatible with their PW counterpart: segments on which the graphs \textit{disagree on sharing} represent only a minor fraction of the genome (much smaller than the fraction of shared or private segments) and their size is compatible with the default block length threshold of PanGraph (100 bp). PM represents and exception, with the great majority of segments being private on each graph and having large size. This expected since for PM the strains considered are highly diverged, above the sensitivity of minimap2 ($\sim 27\%$ core divergence). This causes the majority of blocks not to be merged, and generates a strong order dependence. The size of segments with \textit{disagreement on sharing} is also the largest, some segments having size of 10s of kbp.

\begin{figure}[htb]
    \includegraphics[width=.8\textwidth]{figs_suppl/proj_fig.pdf}
    \caption{{\bf Marginalized graphs are compatible with pairwise graphs}. Comparison of marginalized (MG) and pairwise (PW) built graphs for 5 different species. For each pair of strains we measure the length of segments in which MG and PW graphs agree on whether the segments are shared on both strains (\textit{agree on sharing}). These can either be both shared (\textit{shared on both}) or both private (\textit{private on both}). We also measure the length of segments on which the MG and PW graphs disagree (\textit{disagree on sharing}). For each of these for cases we report the total length as fraction of the average genome (left column) and the average size of each segment (right column). For each category we display the distribution of average results for 50 different pairs (45 for PM).
    }
    \label{fig:marginalize-suppl}
\end{figure}
